Density-dependent selection in evolutionary genetics: a lottery model of Grime's triangle

### 2016-10-31 manuscript.tex
latest

Our mean field approximation is similar to the high-$l_i$ approximation leading to Eq. \eqref{eq:lottery} in that we replace the $x_i$ with appropriate mean values. The key distinction is that territories with a single propagule from the focal genotype, which are critical at low densities, are handled separately. In place of the requirement of $l_i\gg 1$ for all $i$, our approximation only requires that there are no large discrepancies in competitive ability (discussed further below). We obtain (details in Appendix B)
\begin{equation}
\Delta_+ n_i(t)\approx b_i n_i\left[e^{-L}+(R_i+A_i)\frac{c_i}{\overline{c}}\right], \label{eq:master}
\end{equation}
where
\begin{equation}
R_i=\frac{\overline{c}e^{-l_i}(1-e^{-(L-l_i)})}{c_i +\frac{L-1+e^{-L}}{1-(1+L)e^{-L}}\frac{\overline{c}L- c_il_i}{L-l_i}},\label{eq:Dr}
\end{equation}
and
\begin{equation}
A_i=\frac{\overline{c}(1-e^{-l_i})}{c_il_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}+\sum_{j\neq i}\frac{c_jl_j}{L-l_j}\left(L\frac{1-e^{-L}}{1-(1+L)e^{-L}}-l_j\frac{1-e^{-l_j}}{1-(1+l_j)e^{-l_j}}\right)}.\label{eq:Da}
\end{equation}

@@

Fig. \ref{fig:simcomp} shows that Eq. \eqref{eq:master} (and its components) closely approximate direct simulations of random dispersal and lottery competition over a wide range of propagule densities (obtained by varying $U$). Two genotypes are present, one of which has a $c$-advantage and is at low frequency. The growth of the low-frequency genotype relies crucially on the low-density competition term $R_i c_i/\overline{c}$, and also to a lesser extent on the high density competition term $A_i c_i/\overline{c}$ if $l_1$ is large enough (Fig. \ref{fig:simcomp}b). On the other hand, $R_i c_i/\overline{c}$ is negligible for the high-frequency genotype, which depends instead on high density territorial victories (Fig. \ref{fig:simcomp}d).

@@

Combining Eqs. \eqref{eq:equil} and \eqref{eq:invad}, it is easily verified that if $j$ is superior in one trait, but otherwise identical to $i$, it will invade. Moreover, $j$ will eventually exclude $i$, since it is strictly superior. However, stable coexistence is possible between genotypes that are superior in different traits. To illustrate, suppose that $j$ is better at securing territories ($c_j>c_i$), that $i$ is better at producing propagules ($b_i>b_j$), and that $d_i=d_j$. Coexistence occurs if $j$ will invade an $i$-dominated population, but $i$ will also invade a $j$-dominated population (``mutual invasion''). It is not hard to show that this is possible, since if $b_i$ is so large that $L\gg 1$ when $i$ is dominant, and $b_j$ is so small that $L\ll 1$ when $j$ is dominant, then, combining Eqs. \eqref{eq:equil} and \eqref{eq:invad}, we find that $i$ invades $j$ because $b_i>b_j$, while $j$ invades $i$ provided that
\begin{equation}
b_jc_jR_j-b_i c_i A_i>0. \label{eq:jinvadcoex}
\end{equation}
Thus, coexistence occurs if $c_j$ is large enough. Intuitively, the mechanism for coexistence is that territorial contests are important in an $i$-dominated population (high $L$), ensuring that the $c$-specialist $j$ is not excluded, yet territorial contests are irrelevant in a $j$-dominated population (low $L$), ensuring that the $b$-specialist $i$ is not excluded. Fig. \ref{fig:coex} shows an example of this coexistence between $b$ and $c$ specialists.

@@

The mapping of environmental achetypes to our model parameters is summarized in the first two rows of Fig. \ref{fig:table}. Also shown is the approximate dependence of $\Delta_+ n_i$ on $b_i$ and $c_i$ for each archetype (third row). These can be used infer the expected direction of evolution for the traits $b$, $c$ and $d$ (fourth row) as follows.

@@

Following Grime's original argument for a triangular scheme \citep{grime_1977}, Fig. \ref{fig:axes} represents each environmental archetype schematically as a vertex on a triangular space defined by perpendicular stress and disturbance axes. The ideal archetype lies at the origin (no stress or disturbance), while the stress and disturbance archetypes lie at the limits of survival on their respective axes. The hypotenuse connecting the stress and disturbance endpoints represents the limits of survival in the presence of a combination of stress and disturbance. The direction of evolutionary change is different at each vertex, leading to the emergence of different trait clusters or ``primary strategies''.

@@

How does Fig. 4 compare to Grime's C/S/R strategies? In our comparison we will stick to fishes, corals and plants, for which three-way primary strategy schemes are well developed \citep{grime_1977,winemiller_1992,darling_2012}. The connection of our model to fish strategies is necessarily more tentative, given that fishes are motile and not all territorial and the starting assumption of our model is site-occupancy .

@@

Unlike Grime's classic ternary plot \citep{grime_1974}, which represents anti-correlations between traits relevant for success in different environmental archetypes, our realization of Grime's triangle (Fig. 4) refers instead to the direction of trait evolution under different regimes of stress and disturbance. As discussed in section ``Primary strategies and Grime's triangle'', over time our predicted trait evolution should lead to trait values consistent with Grime's scheme. In making these predictions, we have made no reference to any kind of trade-offs or pleitropy, even though trade-offs are often invoked to explain primary strategy schemes \citep{macarthur_1962,winemiller_1992,aerts_1999}. Thus, while trade-offs may amplify specialization, they are not necessary for it. For example, corals which rapidly out-shade neighbors have a tall, branched morphology which is vulnerable to disturbances and so, all else being equal, ideal environment $c$-strategists will suffer higher mortality from disturbances. Yet Fig. \ref{fig:axes} gives the same conclusion; mutations which change disturbance vulnerability are essentially neutral under ideal conditions, whereas higher $c$ is strongly favored.

Our prediction of evolution for higher $b$ in ideal environments is contrary to the expectations of MacArthur's $r$/$K$ dichotomy \citep{macarthur_1962} since $b$ is closely related to the maximal, low-density growth rate $b-d$, and ideal environments support high population densities which should be subject to ``$K$-selection''. However, in the Introduction, we noted that the $r$-$K$ dichotomy is not consistent with empirical studies showing that maximal growth rate and saturation density are positively correlated, both between species/strains \citep{luckinbill_1979,kuno_1991,hendriks_2005,fitzsimmons_2010}, and as a result of experimental evolution \citep{luckinbill_1978,luckinbill_1979}. From the perspective of our model, these correlations are not surprising, since the saturation density, which is determined by a balance between births and deaths, increases with $b$. In reality there may be trade-offs between $b$ and $c$ e.g. due to body size, which we have not accounted for, so that higher $b$ correlates with lower overall competitive success \cite{luckinbill_1979}. Our higher-$b$ predictions simply reflects the fact that, all else being equal, producing more propagules is always advantageous, regardless of population density, a fact lost in the simple logistic interpretation of the $r$/$K$ scheme.

The importance of $b$ at all densities is a general feature of lottery competition. Indeed, as can be seen from Eq. \eqref{eq:lottery}, in the classic lottery model $b_i$ and $c_i$ are essentially equivalent in that only the products $b_i c_i$ matter \citep{chesson_1981}. This is no longer the case in our density-dependent generalization of the classic lottery model, where stable co-existence is possible between $b$ and $c$ strategists. This may seem surprising, given that the classic lottery model was specifically developed for studying species co-existence questions, but the focus in that case was on the role of environmental fluctuations rather than stable coexistence \citep{chesson_1981}. It is not clear whether correctly accounting for the behavior of species at low density will significantly alter the conclusions of \cite{chesson_1981}, but, in any case, species co-existence questions are well beyond the scope of this manuscript. The foremost advantage of our generalized lottery model is that it can account for the growth of rare mutants, making it possible to study evolution in the context of lottery competition. In particular, we are primarily interested in conspecific genotypes rather than the interactions between different species, but there is nothing preventing our model from

Actual K selection

Significance of stage structure

caveats: large c discrepancy

@@

We separate the right hand side of Eq. \eqref{eq:growthsumuncoupled} into three components $\Delta_+ n_i = \Delta_u n_i+\Delta_r n_i+\Delta_a n_i$ which vary in relative magnitude depending on the propagule densities $l_i$. Following the notation in the main text, the Poisson distributions for the $x_i$ (or some subset of the $x_i$) will be denoted $p$; for instance $p(x_1,\ldots,x_G)=p_1(x_1)\ldots p_G(x_G)$ and $p(x_1,\ldots,x_{i-1},x_{i+1}\ldots,x_G)=p_1(x_1)\ldots p_{i-1}(x_{i-1})p_{i+1}(x_{i+1})\ldots p_G(x_G)$. We use $P$ as a general shorthand for the probability of particular outcomes.

@@

We now show that, with respect to $\tilde{p}$, the standard deviation in $\sum_{j\neq i} c_j x_j$, $\sigma(\sum_{j\neq i} c_j x_j)$, is much smaller than its mean $\langle\sum_{j\neq i} c_j x_j\rangle_{\tilde{p}}$. Then $x_j$ can be replaced by its mean in the last term in Eq. \eqref{eq:deltr},
\begin{equation}
\left\langle\frac{c_i}{c_i +\sum_{j\neq i} c_j x_j}\right\rangle_{\tilde{p}}\approx \frac{c_i}{c_i +\sum_{j\neq i} c_j \langle x_j\rangle_{\tilde{p}}},\label{eq:meanfieldr}
\end{equation}
which will give us Eq. \eqref{eq:Dr}.

@@

To calculate the relative fluctuations in $\sum_{j\neq i} c_j x_j$, we use the following approximation, which gives considerably simpler expressions for the means, variances and covariances of the $x_j$ compared with the exact expressions using $\tilde{p}$. Rather than evaluating the situation in each territory after dispersal as above, we let $\tilde{p}$ instead be the ${\mathbf x_i}$ dispersal probabilities in a territory where one focal propagule is assumed to be present, conditional on $\sum_{j\neq i} x_j>1$. This gives $\langle x_j \rangle_{\tilde{p}}=l_j/C$, 
\begin{equation}
\sigma^2(x_j)=\frac{l_j^2}{C}\left(1-\frac{1}{C}\right)+\frac{l_j}{C},\label{eq:varr}
\end{equation}
and 
\begin{equation}
\sigma(x_j,x_k)=\frac{l_j l_k}{C}\left(1-\frac{1}{C}\right),\label{eq:covr}
\end{equation}
where $C=1-e^{-(L-l_i)}$ (note the difference from Eq. \eqref{eq:meanxjrare} for $\langle x_j \rangle_{\tilde{p}}$). Then, since
\begin{equation}
\sigma^2(\sum_{j\neq i} c_j x_j)=\sum_{j\neq i}\left[c_j^2\sigma^2(x_j)+2\sum_{k>j}c_j c_k\sigma(x_j,x_k)\right],\label{eq:vartotr}
\end{equation}
and $1/C>1$, we have
\begin{equation}
\frac{\sigma(\sum_{j\neq i} c_j x_j)}{\langle\sum_{j\neq i} c_j x_j\rangle}<C^{1/2}\frac{\left(\sum_{j\neq i}c_j^2 l_j\right)^{1/2}}{\sum_{j\neq i}c_j l_j}. \label{eq:cvr}
\end{equation}

@@

When at least some of the nonfocal propagule densities are large $l_j\gg 1$, then the RHS of Eq. \eqref{eq:cvr} is $\ll 1$, as desired. This is also the case if none of the nonfocal genotype densities are large and the $c_j$ are all of similar magnitude (their ratios are of order one); the worst case scenario occurs when $(L-l_i)\sim O(1)$, in which case the negative covariances (Eq. \eqref{eq:covr}) which were neglected in the RHS of Eq. \eqref{eq:cvr} significantly reduce the overall variance $\sigma^2(\sum_{j\neq i} c_j x_j)$.

However, the relative fluctuations in $\sum_{j\neq i} c_j x_j$ can be large if some of the $c_j$ are much larger than the others. Specifically, if $c_j l_j\gg c_k l_k$ ($j,k\neq i$, $j\neq k$) and $l_j\ll 1$ (i.e. in the presence of a rare, extremely strong competitor), then we cannot make the replacement Eq. \eqref{eq:meanfieldr}.

Substituting Eqs. \eqref{eq:meanfieldr} and \eqref{eq:meanxjrare} into Eq. \eqref{eq:deltr}, we obtain
\begin{equation}
\Delta_r n_i\approx m_i R_i\frac{c_i}{\overline{c}}, \label{eq:deltrfinal}
\end{equation}
where $R_i$ is defined in Eq. \eqref{eq:Dr}.

@@

The final contribution, $\Delta_a n_i$, accounts for territories where two or more focal propagules are present ($a$ stands for ``abundant"). Similarly to Eq. \eqref{eq:deltr}, we have 
\begin{equation}
\Delta_a n_i=U(1-(1+l_i)e^{l_i})\left\langle \frac{c_i x_i}{\sum_j c_j x_j} \right\rangle_{\hat{p}}\label{eq:delta}
\end{equation}
where $\hat{p}$ is the probability distribution of both focal and nonfocal propagaule abundances \textit{after} dispersal in those territories where at least two focal propagules landed.

@@

To calculate the relative fluctuations in $\sum_{j\neq i} c_j x_j$, we use a similar approximation as for $\Delta_r n_i$: $\hat{p}$ is approximated by the ${\mathbf x}$ dispersal probabilities in a territory where at least two focal propagule is assumed to be present. All covariances are now zero, so that $\sigma^2(\sum c_j x_j)=\sum c_j^2 \sigma^2(x_j)$, where $\sigma^2(x_j)=l_j$ for $j\neq i$. The expression for $\sigma^2(x_i)$ is more complicated. We assume $p(x_i=0)\approx 0$ without loss of generality (since otherwise $D\gg 1$ and $\Delta n_a$ is negligible). Then
\begin{equation}
\sigma^2(x_i)=\frac{l_i^2}{D}\left(1-\frac{1}{D}\right)+\frac{l_i}{D},
\end{equation}
where $D=1-(1+l_i)e^{-l_i}$, analogous to Eq. \eqref{eq:varr}, and 
\begin{equation}
\frac{\sigma(\sum c_j x_j)}{\langle\sum c_j x_j\rangle} \approx\frac{\left(\sum_{j\neq i} c_j^2 l_j + c_i^2 \sigma^2(x_i)\right)^{1/2}}{\sum_{j\neq i} c_j l_j + c_i l_i/D} \label{eq:cva}.
\end{equation}

Similarly to Eq. \eqref{eq:cvr}, the RHS of \eqref{eq:cva} will not be $\ll1$ if there is a nonfocal genotype $j$ with $l_j\ll 1$ and $c_j l_j\gg c_k l_k$ for $j,k\neq i$, $j\neq k$. When this is not the case, then since $l_i$ must be of order $1$ or larger for $\Delta_a n$ to make an appreciable contribution to $\Delta_+ n_i$, the RHS of Eq. \eqref{eq:cva} is $\ll 1$ as desired.

Combining Eqs. \eqref{eq:delta} and \eqref{eq:meanfielda}, we obtain
\begin{equation}
\Delta_a n_i=m_i A_i \frac{c_i}{\overline{c}},
\end{equation}
where $A_i$ is defined in Eq. \eqref{eq:Da}.

%The logistic equation for a single genotype or species is given by
%\begin{equation}
%\frac{dN}{dt}=r\left(1-\frac{N}{K}\right)N \label{eq:logistic}
%\end{equation}
%where $N$ is the total population abundance or biomass, $r$ is the growth rate at  low densities, and the ``carrying capacity" $K$ is the value of $N$ where growth ceases.
%
%a simple, ubiquitous model of density-dependent population growth
%declines linearly with density,
%
%Eq. \eqref{eq:logistic} is often interpreted as a linear approximation of a more complicated growth equation
%\begin{equation}
%\frac{dN}{dt}=r(N)N,
%\end{equation}
%where the growth rate $r(N)$ is some unknown, possibly complicated, function of $N$.

>>>>

Our mean field approximation is similar to the high-$l_i$ approximation leading to Eq. \eqref{eq:lottery} in that we replace the $x_i$ with appropriate mean values. The key distinction is that territories with a single propagule from the focal genotype, which are critical at low densities, are handled separately. In place of the requirement of $l_i\gg 1$ for all $i$, our approximation only requires that there are no large discrepancies in competitive ability (specifically, that we do not have $c_i/c_j\gg 1$ for any two genotypes; further discussion in section ``Discussion''). We obtain (details in Appendix B)
\begin{equation}
\Delta_+ n_i(t)\approx b_i n_i\left[e^{-L}+(R_i+A_i)\frac{c_i}{\overline{c}}\right], \label{eq:master}
\end{equation}
where
\begin{equation}
R_i=\frac{\overline{c}e^{-l_i}(1-e^{-(L-l_i)})}{c_i +\frac{L-1+e^{-L}}{1-(1+L)e^{-L}}\frac{\overline{c}L- c_il_i}{L-l_i}},\label{eq:Dr}
\end{equation}
and
\begin{equation}
A_i=\frac{\overline{c}(1-e^{-l_i})}{c_il_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}+\sum_{j\neq i}\frac{c_jl_j}{L-l_j}\left(L\frac{1-e^{-L}}{1-(1+L)e^{-L}}-l_j\frac{1-e^{-l_j}}{1-(1+l_j)e^{-l_j}}\right)}.\label{eq:Da}
\end{equation}

@@

Fig.~\ref{fig:simcomp} shows that Eq. \eqref{eq:master} (and its components) closely approximate direct simulations of random dispersal and lottery competition over a wide range of propagule densities (obtained by varying $U$). Two genotypes are present, one of which has a $c$-advantage and is at low frequency. The growth of the low-frequency genotype relies crucially on the low-density competition term $R_i c_i/\overline{c}$, and also to a lesser extent on the high density competition term $A_i c_i/\overline{c}$ if $l_1$ is large enough (Fig.~\ref{fig:simcomp}b). On the other hand, $R_i c_i/\overline{c}$ is negligible for the high-frequency genotype, which depends instead on high density territorial victories (Fig.~\ref{fig:simcomp}d).

@@

Combining Eqs. \eqref{eq:equil} and \eqref{eq:invad}, it is easily verified that if $j$ is superior in one trait, but otherwise identical to $i$, it will invade. Moreover, $j$ will eventually exclude $i$, since it is strictly superior. However, stable coexistence is possible between genotypes that are superior in different traits. To illustrate, suppose that $j$ is better at securing territories ($c_j>c_i$), that $i$ is better at producing propagules ($b_i>b_j$), and that $d_i=d_j$. Coexistence occurs if $j$ will invade an $i$-dominated population, but $i$ will also invade a $j$-dominated population (``mutual invasion''). It is not hard to show that this is possible, since if $b_i$ is so large that $L\gg 1$ when $i$ is dominant, and $b_j$ is so small that $L\ll 1$ when $j$ is dominant, then, combining Eqs. \eqref{eq:equil} and \eqref{eq:invad}, we find that $i$ invades $j$ because $b_i>b_j$, while $j$ invades $i$ provided that
\begin{equation}
b_jc_jR_j-b_i c_i A_i>0. \label{eq:jinvadcoex}
\end{equation}
Thus, coexistence occurs if $c_j$ is large enough. Intuitively, the mechanism for coexistence is that territorial contests are important in an $i$-dominated population (high $L$), ensuring that the $c$-specialist $j$ is not excluded, yet territorial contests are irrelevant in a $j$-dominated population (low $L$), ensuring that the $b$-specialist $i$ is not excluded. Fig.~\ref{fig:coex} shows an example of this coexistence between $b$ and $c$ specialists.

@@

The mapping of environmental achetypes to our model parameters is summarized in the first two rows of Fig.~\ref{fig:table}. Also shown is the approximate dependence of $\Delta_+ n_i$ on $b_i$ and $c_i$ for each archetype (third row). These can be used infer the expected direction of evolution for the traits $b$, $c$ and $d$ (fourth row) as follows.

@@

Following Grime's original argument for a triangular scheme \citep{grime_1977}, Fig.~\ref{fig:axes} represents each environmental archetype schematically as a vertex on a triangular space defined by perpendicular stress and disturbance axes. The ideal archetype lies at the origin (no stress or disturbance), while the stress and disturbance archetypes lie at the limits of survival on their respective axes. The hypotenuse connecting the stress and disturbance endpoints represents the limits of survival in the presence of a combination of stress and disturbance. The direction of evolutionary change is different at each vertex, leading to the emergence of different trait clusters or ``primary strategies''.

@@

How does Fig.~4 compare to Grime's C/S/R strategies? In our comparison we will stick to fishes, corals and plants, for which three-way primary strategy schemes are well developed \citep{grime_1977,winemiller_1992,darling_2012}. The connection of our model to fish strategies is necessarily more tentative, given that fishes are motile and not all territorial and the starting assumption of our model is site-occupancy .

@@

Unlike Grime's classic ternary plot \citep{grime_1974}, which represents anti-correlations between traits relevant for success in different environmental archetypes, our realization of Grime's triangle (Fig.~4) refers instead to the direction of trait evolution under different regimes of stress and disturbance. As discussed in section ``Primary strategies and Grime's triangle'', over time our predicted trait evolution should lead to trait values consistent with Grime's scheme. In making these predictions, we have made no reference to any kind of trade-offs or pleitropy, even though trade-offs are often invoked to explain primary strategy schemes \citep{macarthur_1962,winemiller_1992,aerts_1999}. Thus, while trade-offs may amplify specialization, they are not necessary for it. For example, corals which rapidly out-shade neighbors have a tall, branched morphology which is vulnerable to disturbances, and so, all else being equal, ideal environment $c$-strategists will suffer higher mortality from disturbances. Yet Fig.~\ref{fig:axes} gives the same conclusion without invoking trade-offs; mutations which reduce disturbance vulnerability are essentially neutral under ideal conditions, leading to no improvements in mortality from disturbances, whereas $c$ will tend to increase over time.

Our prediction of evolution for higher $b$ in ideal environments is contrary to the expectations of MacArthur's $r$/$K$ dichotomy \citep{macarthur_1962} since $b$ is closely related to the maximal, low-density growth rate $b-d$, and ideal environments support high population densities which should be subject to ``$K$-selection''. However, in the Introduction, we noted that the $r$-$K$ dichotomy is not consistent with empirical studies showing that maximal growth rate and saturation density are positively correlated, both between species/strains \citep{luckinbill_1979,kuno_1991,hendriks_2005,fitzsimmons_2010}, and as a result of experimental evolution \citep{luckinbill_1978,luckinbill_1979}. From the perspective of our model, these correlations are not surprising, since the saturation density, which is determined by a balance between births and deaths, increases with $b$. Our higher-$b$ predictions simply reflects the fact that, all else being equal, producing more propagules is always advantageous, regardless of population density, a fact lost in the simple logistic interpretation of the $r$/$K$ scheme.

Confusingly, the term ``$K$-selection'' has sometimes been used to refer generally to selection at high density; this encompasses both selection for higher saturation density --- ``$K$'' in the logistic equation --- as well as selection for competitive ability. The latter form of selection has been termed $\alpha$-selection after the competition coefficients in the Lotka-Volterra equation \citep{gill_1974,case_1974,joshi_2001}. Unlike the correlation

In contrast to the problems with a strict $r$/$K$ dichotomy, there are

In reality there may be trade-offs between $b$ and $c$ e.g. due to body size, which we have not accounted for, so that higher $b$ correlates with lower overall competitive success \cite{luckinbill_1979}.

Actual K selection,

Lotka Volterra with alpha,

resource comp.

significance of stage structure

The importance of $b$ in securing territories is a general feature of lottery competition. Indeed, as can be seen from Eq. \eqref{eq:lottery}, in the classic lottery model $b_i$ and $c_i$ are essentially equivalent in that only the products $b_i c_i$ matter \citep{chesson_1981}. This is no longer the case in our density-dependent generalization of the classic lottery model, where stable co-existence is possible between $b$ and $c$ strategists. Given that the classic lottery model was specifically developed for studying species co-existence questions, this may seem surprising, but the focus in that case was on the role of environmental fluctuations in promoting co-existence rather than stable coexistence \citep{chesson_1981}. It is not clear whether correctly accounting for the behavior of species with low density would significantly alter the conclusions of \cite{chesson_1981}, but, in any case, species co-existence questions are beyond the scope of this manuscript.

More broadly, while our model can be applied to questions at a species level (e.g. ecological invasions), our focus here is on the evolution of genotype frequencies within a population. This is only possible because the model accounts for the growth of mutants from low densities. Given this focus, our assumption that there are no large $c$ discrepancies (section ``Mean field approximation'') amounts to a restriction on the amount of genetic variation in $c$ that will be sustained in the population. Since beneficial mutation effect sizes will typically not be much larger than a few percent, large $c$ discrepancies can only arise if the mutation rate is extremely large, and so the assumption will not be violated in most cases. However, more care would be needed if we were looking species interactions instead of genotype evolution.

eco-evo stuff

@@

We separate the right hand side of Eq.~\eqref{eq:growthsumuncoupled} into three components $\Delta_+ n_i = \Delta_u n_i+\Delta_r n_i+\Delta_a n_i$ which vary in relative magnitude depending on the propagule densities $l_i$. Following the notation in the main text, the Poisson distributions for the $x_i$ (or some subset of the $x_i$) will be denoted $p$; for instance $p(x_1,\ldots,x_G)=p_1(x_1)\ldots p_G(x_G)$ and $p(x_1,\ldots,x_{i-1},x_{i+1}\ldots,x_G)=p_1(x_1)\ldots p_{i-1}(x_{i-1})p_{i+1}(x_{i+1})\ldots p_G(x_G)$. We use $P$ as a general shorthand for the probability of particular outcomes.

@@

We now show that, with respect to $\tilde{p}$, the standard deviation in $\sum_{j\neq i} c_j x_j$, $\sigma(\sum_{j\neq i} c_j x_j)$, is much smaller than its mean $\langle\sum_{j\neq i} c_j x_j\rangle_{\tilde{p}}$. Then $x_j$ can be replaced by its mean in the last term in Eq.~\eqref{eq:deltr},
\begin{equation}
\left\langle\frac{c_i}{c_i +\sum_{j\neq i} c_j x_j}\right\rangle_{\tilde{p}}\approx \frac{c_i}{c_i +\sum_{j\neq i} c_j \langle x_j\rangle_{\tilde{p}}},\label{eq:meanfieldr}
\end{equation}
which will give us Eq.~\eqref{eq:Dr}.

@@

To calculate the relative fluctuations in $\sum_{j\neq i} c_j x_j$, we use the following approximation, which gives considerably simpler expressions for the means, variances and covariances of the $x_j$ compared with the exact expressions using $\tilde{p}$. Rather than evaluating the situation in each territory after dispersal as above, we let $\tilde{p}$ instead be the ${\mathbf x_i}$ dispersal probabilities in a territory where one focal propagule is assumed to be present, conditional on $\sum_{j\neq i} x_j>1$. This gives $\langle x_j \rangle_{\tilde{p}}=l_j/C$, 
\begin{equation}
\sigma^2(x_j)=\frac{l_j^2}{C}\left(1-\frac{1}{C}\right)+\frac{l_j}{C},\label{eq:varr}
\end{equation}
and 
\begin{equation}
\sigma(x_j,x_k)=\frac{l_j l_k}{C}\left(1-\frac{1}{C}\right),\label{eq:covr}
\end{equation}
where $C=1-e^{-(L-l_i)}$ (note the difference from Eq.~\eqref{eq:meanxjrare} for $\langle x_j \rangle_{\tilde{p}}$). Then, since
\begin{equation}
\sigma^2(\sum_{j\neq i} c_j x_j)=\sum_{j\neq i}\left[c_j^2\sigma^2(x_j)+2\sum_{k>j}c_j c_k\sigma(x_j,x_k)\right],\label{eq:vartotr}
\end{equation}
and $1/C>1$, we have
\begin{equation}
\frac{\sigma(\sum_{j\neq i} c_j x_j)}{\langle\sum_{j\neq i} c_j x_j\rangle}<C^{1/2}\frac{\left(\sum_{j\neq i}c_j^2 l_j\right)^{1/2}}{\sum_{j\neq i}c_j l_j}. \label{eq:cvr}
\end{equation}

@@

When at least some of the nonfocal propagule densities are large $l_j\gg 1$, then the RHS of Eq.~\eqref{eq:cvr} is $\ll 1$, as desired. This is also the case if none of the nonfocal genotype densities are large and the $c_j$ are all of similar magnitude (their ratios are of order one); the worst case scenario occurs when $(L-l_i)\sim O(1)$, in which case the negative covariances (Eq.~\eqref{eq:covr}) which were neglected in the RHS of Eq.~\eqref{eq:cvr} significantly reduce the overall variance $\sigma^2(\sum_{j\neq i} c_j x_j)$.

However, the relative fluctuations in $\sum_{j\neq i} c_j x_j$ can be large if some of the $c_j$ are much larger than the others. Specifically, if $c_j l_j\gg c_k l_k$ ($j,k\neq i$, $j\neq k$) and $l_j\ll 1$ (i.e. in the presence of a rare, extremely strong competitor), then we cannot make the replacement Eq.~\eqref{eq:meanfieldr}.

Substituting Eqs. \eqref{eq:meanfieldr} and \eqref{eq:meanxjrare} into Eq.~\eqref{eq:deltr}, we obtain
\begin{equation}
\Delta_r n_i\approx m_i R_i\frac{c_i}{\overline{c}}, \label{eq:deltrfinal}
\end{equation}
where $R_i$ is defined in Eq.~\eqref{eq:Dr}.

@@

The final contribution, $\Delta_a n_i$, accounts for territories where two or more focal propagules are present ($a$ stands for ``abundant"). Similarly to Eq.~\eqref{eq:deltr}, we have 
\begin{equation}
\Delta_a n_i=U(1-(1+l_i)e^{l_i})\left\langle \frac{c_i x_i}{\sum_j c_j x_j} \right\rangle_{\hat{p}}\label{eq:delta}
\end{equation}
where $\hat{p}$ is the probability distribution of both focal and nonfocal propagaule abundances \textit{after} dispersal in those territories where at least two focal propagules landed.

@@

To calculate the relative fluctuations in $\sum_{j\neq i} c_j x_j$, we use a similar approximation as for $\Delta_r n_i$: $\hat{p}$ is approximated by the ${\mathbf x}$ dispersal probabilities in a territory where at least two focal propagule is assumed to be present. All covariances are now zero, so that $\sigma^2(\sum c_j x_j)=\sum c_j^2 \sigma^2(x_j)$, where $\sigma^2(x_j)=l_j$ for $j\neq i$. The expression for $\sigma^2(x_i)$ is more complicated. We assume $p(x_i=0)\approx 0$ without loss of generality (since otherwise $D\gg 1$ and $\Delta n_a$ is negligible). Then
\begin{equation}
\sigma^2(x_i)=\frac{l_i^2}{D}\left(1-\frac{1}{D}\right)+\frac{l_i}{D},
\end{equation}
where $D=1-(1+l_i)e^{-l_i}$, analogous to Eq.~\eqref{eq:varr}, and 
\begin{equation}
\frac{\sigma(\sum c_j x_j)}{\langle\sum c_j x_j\rangle} \approx\frac{\left(\sum_{j\neq i} c_j^2 l_j + c_i^2 \sigma^2(x_i)\right)^{1/2}}{\sum_{j\neq i} c_j l_j + c_i l_i/D} \label{eq:cva}.
\end{equation}

Similarly to Eq.~\eqref{eq:cvr}, the RHS of \eqref{eq:cva} will not be $\ll1$ if there is a nonfocal genotype $j$ with $l_j\ll 1$ and $c_j l_j\gg c_k l_k$ for $j,k\neq i$, $j\neq k$. When this is not the case, then since $l_i$ must be of order $1$ or larger for $\Delta_a n$ to make an appreciable contribution to $\Delta_+ n_i$, the RHS of Eq.~\eqref{eq:cva} is $\ll 1$ as desired.

Combining Eqs. \eqref{eq:delta} and \eqref{eq:meanfielda}, we obtain
\begin{equation}
\Delta_a n_i=m_i A_i \frac{c_i}{\overline{c}},
\end{equation}
where $A_i$ is defined in Eq.~\eqref{eq:Da}.

%The logistic equation for a single genotype or species is given by
%\begin{equation}
%\frac{dN}{dt}=r\left(1-\frac{N}{K}\right)N \label{eq:logistic}
%\end{equation}
%where $N$ is the total population abundance or biomass, $r$ is the growth rate at  low densities, and the ``carrying capacity" $K$ is the value of $N$ where growth ceases.
%
%a simple, ubiquitous model of density-dependent population growth
%declines linearly with density,
%
%Eq.~\eqref{eq:logistic} is often interpreted as a linear approximation of a more complicated growth equation
%\begin{equation}
%\frac{dN}{dt}=r(N)N,
%\end{equation}
%where the growth rate $r(N)$ is some unknown, possibly complicated, function of $N$.
